A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-13 paper.tex
Incorporating suggestions

\texttt{KwARG} \citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination breakpoints, which has been shown to be the true minimum required to explain the dataset when disallowing recurrent mutation \citep{song2003parsimonious}. An ARG drawn from the posterior sample produced by \texttt{ARGweaver} has 17 breakpoints. Both tools explicitly record the sequence undergoing each recombination event, and the precise location of the breakpoint along the genome. \texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} instead construct a sequence of correlated local trees, but without specifying the precise positions of recombination events (i.e.\ reconstructing the local trees without explicitly specifying the SPR operations required to transition between them). The number of inferred trees also depends on the parameter settings; all four tools can allow for the presence of sequencing errors and recurrent mutations, reducing the number of inferred local trees (in this case, based on the input mutation and recombination rate \texttt{Relate} hence posits only two recombination breakpoints).

There is some consistency among the four inferred ARGs (such as the samples FR-F, WA-F and Af-F being grouped into the same clade), but also considerable differences in the order and number of events. Note that all four ARGs can be represented in the succinct tree sequence format, with the ARGs produced by \texttt{KwARG} and \texttt{ARGweaver} containing more nodes to record the recombination events (corresponding to Figure 4 stage `B'), and the ARGs output by \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (Figure 4 stage `D').

>>>>

There is some consistency among the four inferred ARGs (such as the samples FR-F, WA-F and Af-F being grouped into the same clade), but also considerable differences in the order and number of events. \texttt{KwARG} \citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination breakpoints, which has been shown to be the true minimum required to explain the dataset when disallowing recurrent mutation \citep{song2003parsimonious}; this is a lower bound that, in general, significantly underestimates the actual number of recombinations that might have occurred. An ARG drawn from the posterior sample produced by \texttt{ARGweaver} has 17 breakpoints. Both tools explicitly record the sequence undergoing each recombination event, and the precise location of the breakpoint along the genome. \texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} do not place recombination events onto the ARG topology, therefore the transitions between local trees (and corresponding SPR operations) are not explicitly inferred. \texttt{Relate} first infers local tree topologies and then identifies edges that persist from one tree to the next as a separate step, while \texttt{tsinfer} explicitly constructs nodes that are shared across local trees. The number of inferred trees also depends on the parameter settings; all four tools can allow for the presence of sequencing errors and recurrent mutations, reducing the number of inferred local trees (in this case, based on the input mutation and recombination rate \texttt{Relate} hence posits only two recombination breakpoints).

Note that all four methods produce output which can be represented in the succinct tree sequence format, and which is consistent with our definition of an ARG as a data structure. The ARGs produced by \texttt{KwARG} and \texttt{ARGweaver} contain more nodes to record the recombination events (corresponding to Figure 4 stage `B'), while the ARGs output by \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (roughly corresponding to Figure 4 stages `C' and 'D', respectively).
